Type theory in type theory with single substitutions

### 2025-01-30 cpp/p.tex
stuff

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory more abstractly than GATs: SOGATs abstract over the exact
definition of the substitution calculus (whether substitutions are
single or parallel; whether we have CwF-style or Ehrhard-style
operations): contexts, substitutions, instantiations are simply
modelled by the metatheoretic function space. For example, a type
theory with $\Pi$, Coquand-universes and a universe level lifting
operation is described by the following SOGAT (where $\cong$ means
isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved, as there is no
good notion of homomorphism between them. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
SOGAT then is a model of the GAT the we obtain by their
translation. GATs come with well-behaved metatheory: a category of
models witha initial model (syntax), (co)free models
\cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
translation is not unique: Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} define two different translations, one
based on parallel substitutions and one based on single
substitutions. If we apply their parallel translation to the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we use their single substitution
translation to the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). The current paper
can be seen as taking a big enough type theory (described as a SOGAT),
and investigating how the result of its single substitution
translation relates to the result of its parallel translation (the
latter first-order theory is quite well-known). There is no general
result about the relationship of the two translations.

@@

In this section, we introduce the syntax of type theory with function
space and universes in a minimalistic way, only introducing operations
that are unavoidable. We eschew boilerplate by only defining
well-formed, well-scoped, well-typed terms that are quotiented by
conversion. This means that the equalities that hold between terms are
the ways we can convert terms into each other when running them as
programs. We give a tutorial-style introduction to the syntax, we do
not assume prior knowledge of the metatheory of type theory. The
definition of type theory that we obtain is the same as the single
substitution GAT semantics \cite{DBLP:conf/fscd/KaposiX24} of the
SOGAT (\ref{eq:tt}).

@@

@@

Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
\begin{alignat*}{10}
  & \lam\,(\lam\,\q)\cdot\cd\,A\cdot a && {=}(\Pi\beta) \\
  & (\lam\,\q[\langle\cd\,A\rangle])\cdot a && {=}(\lam[]) \\
  & (\lam\,(\q[\langle\cd\,A\rangle^+]))\cdot a && {=}([\q][^+]) \\
  & (\lam\,\q)\cdot a && {=}(\Pi\beta) \\
  & \q[\langle a\rangle] && {=}(\q[\langle\rangle]) \\
  & a
\end{alignat*}
If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
$a' : \Tm\,(\diamond\ext C)\,A'$, we have to weaken our function to
accept them:
\begin{alignat*}{10}
  & \lam\,(\lam\,\q)[\p] : \Tm\,&& (\diamond\ext C)\,\\
  & && \Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big)
\end{alignat*}
But we have
\[
\lam\,(\lam\,\q)[\p] \overset{\lam[]\,2\text{x}}{=} \lam\,(\lam\,(\q[\p^{++}])) \overset{\q[^+]}{=} \lam\,(\lam\,\q)
\]
(with different implicit arguments), so we are able apply $\cd\,A'$
and $a'$ just as before. What we cannot do is to directly weaken a
term in context $\diamond$ to an arbitrary metavariable context
$\Gamma$, we can only do this step by step by applying weakening
$\blank[\p]$ for each type in the context.

@@

$\Sigma$ types were introduced by a similar isomorphism, and we give
names to the roundtrips according to our convention introduced in
Section \ref{sec:metatheory}. xxx

but
naturality was only stated for the right to left direction (rule
${,}[]$). Naturalities in the direction are derivable:
\begin{alignat*}{10}
  & (\fst\,w)[\gamma] && {=}(xxx) \\
  & \fst\,((\fst\,w)[\gamma],(\snd\,w)[\gamma]) = \\
  & \fst\,((\fst\,w,\snd\,w)[\gamma]) = \\
  & \fst\,(w[\gamma])
\end{alignat*}

We show the commutation of $\Lift$ and $\Pi$ types as a sequence of
isomorphisms:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,(\Lift\,(\Pi\,A\,B)) && {\cong}\text{(1)} \\
  & \Tm\,\Gamma\,(\Pi\,A\,B) && {\cong}\text{(2)} \\
  & \Tm\,(\Gamma\ext A)\,B && {\cong}\text{(3)} \\
  & \Tm\,(\Gamma\ext A)\,(\Lift\,B) && {\cong}\text{(4)} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) && {\cong}\text{(5)} \\
  & \Tm\,\Gamma\,(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]))
\end{alignat*}
Steps (1) and (3) are the isomorphism comprised of $\mk$ and $\un$.

@@

As a sanity check for our notion of SSC-model, we defined the standard
(metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
SSC-based calculus calculus in Agda. We used the techniques of Kovács
\cite{DBLP:conf/csl/Kovacs22} to define inductive-recursive universes
to model the infinite hierarchy of types, and all the SSC-equations
hold by reflexivity in this model. We defined the analogous CwF-based
standard model.

@@

In this section we show that our SSC-based syntax is isomorphic to the
CwF-based syntax for the same theory (namely the SOGAT
(\ref{eq:tt})). We conjecture that the method presented here works
generically for any SOGAT.

@@

The following four equations correspond to equations in B-systems which cannot
be derived from the equations of SSC, these are versions of equations in Section
\ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
\begin{equation}\label{eq:lifted}
\begin{aligned}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
\end{aligned}
\end{equation}
The corresponding equations on terms are also needed. Note that we do not have
the non-lifted term versions of the last two equations in SSC.

To formally state the equations in (\ref{eq:lifted}), we define telescopes as an
inductive type together with a recursive operation to append telescopes to
contexts.
\begin{alignat*}{10}
  & \Tel && : \Con\ra\Set \\
  & \blank+\blank && : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con \\
  & \diamond && : \Tel\,\Gamma \\
  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Omega
\end{alignat*}
The appending operation is defined recursively on telescopes.
\begin{alignat*}{10}
  & \Gamma+\diamond && := \Gamma \\
  & \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
\end{alignat*}
We then define a lifting operation over any telescope mutually with an
instantiation operation on telescopes by recursion on telescopes.
\begin{alignat*}{10}
  & \blank[\blank] && : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \\
  & \blank^{+^\Omega} && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
\end{alignat*}

If we try to prove the equations in (\ref{eq:lifted}) by induction on the types
and terms, we run into difficulties in the instantiation case, as we would have
to commute the instantiations without the equations to do so. Therefore, we
first $\alpha$-normalize the syntax to compute away all the instantiations, then
prove the equations by induction on the $\alpha$-normal forms.

@@

Instead of doing induction on $\alpha$-normal forms for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules are
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if $A[\gamma_0] =
  A[\gamma_1]$ and $x[\gamma_0] = x[\gamma_1]$ for any $A:\Ty\,\Gamma$ and
  variable $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\gamma_0^{+^\Omega}] =
  x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}

>>>>

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory more abstractly than GATs: SOGATs abstract over the exact
definition of the substitution calculus (whether substitutions are
single or parallel; whether we have CwF-style or Ehrhard-style
operations): contexts, substitutions, instantiations are simply
modelled by the metatheoretic function space. For example, a type
theory with $\Pi$, Coquand-universes and a universe level lifting
operation is described by the following SOGAT (where $\cong$ means
isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved, as there is no
good notion of homomorphism between them. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
SOGAT then is a model of the GAT the we obtain by their
translation. GATs come with well-behaved metatheory: a category of
models with initial model (syntax), (co)free models
\cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
translation is not unique: Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} define two different translations, one
based on parallel substitutions and one based on single
substitutions. If we apply their parallel translation to the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we apply their single substitution
translation to the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). The current paper
can be seen as taking a big enough type theory (described as a SOGAT),
and investigating how the result of its single substitution
translation relates to the result of its parallel translation (the
latter first-order theory is quite well-known). There is no general
result about the relationship of the two translations.

@@

In this section, we introduce the syntax of type theory with function
space and universes in a minimalistic way, only introducing operations
that are unavoidable. We eschew boilerplate by only defining
well-formed, well-scoped, well-typed terms that are quotiented by
conversion. This means that the equalities that hold between terms are
the ways we can convert terms into each other when running them as
programs. We give a tutorial-style introduction to the syntax, we do
not assume prior knowledge of the metatheory of type theory.

@@

In this subsection, we show how to use the above defined calculus, and
motivate the next section on admissible equations.

@@

Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
\begin{alignat*}{10}
  & \lam\,(\lam\,\q)\cdot\cd\,A\cdot a && {=}(\Pi\beta) \\
  & (\lam\,\q[\langle\cd\,A\rangle])\cdot a && {=}(\lam[]) \\
  & (\lam\,(\q[\langle\cd\,A\rangle^+]))\cdot a && {=}([\q][^+]) \\
  & (\lam\,\q)\cdot a && {=}(\Pi\beta) \\
  & \q[\langle a\rangle] && {=}(\q[\langle\rangle]) \\
  & a
\end{alignat*}
If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
$a' : \Tm\,(\diamond\ext C)\,A'$, we have to weaken our function to
accept them:
\begin{alignat*}{10}
  & \lam\,(\lam\,\q)[\p] : \Tm\,&& (\diamond\ext C)\,\\
  & && \Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big)
\end{alignat*}
But we have
\[
\lam\,(\lam\,\q)[\p] \overset{\lam[]\,2\text{x}}{=} \lam\,(\lam\,(\q[\p^{++}])) \overset{\q[^+]}{=} \lam\,(\lam\,\q)
\]
(with different implicit arguments), so we are able apply $\cd\,A'$
and $a'$ just as before. What we cannot do is to directly weaken a
term in context $\diamond$ to an arbitrary context $\Gamma$, where
$\Gamma$ is a metavariable. If we know that $\Gamma$ has lengh $n$,
then we can do the weakening by
$\blank\underbrace{[\p][\p]\dots[\p]}_{\text{$n$ times}}$. If we work
in the syntax, then we can derive such weakenings (and more) using the
methods of Section \ref{sec:admissible}.

@@

$\Sigma$ types were introduced by a similar isomorphism, and we give
names to the roundtrips eqautions according to our convention introduced in
Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
$\Sigma\beta_2 : \snd\,(a,b) = b$ and $\Sigma\eta : w =
(\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
direction (${,}[]$), because naturalities in the other direction are
derivable:
\begin{alignat*}{10}
  & (\fst\,w)[\gamma] && {=}(\Sigma\beta_1)                             && (\snd\,w)[\gamma] && {=}(\Sigma\beta_2)                             \\
  & \fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \hspace{4em} && \snd\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \\
  & \fst\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        && \snd\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        \\
  & \fst\,(w[\gamma])                                                && && \snd\,(w[\gamma])                                                   
\end{alignat*}
For $\Pi$ types, we construct a similar isomorphism:
\[
(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam
\]
The roundtrips are proven as follows:
\[
\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t
\]
The last step is the admissible equation (\ref{eq:lifted4}) proven in
Section \ref{sec:admissible}. As above with weakening into an
arbitrary context, this equation holds in the syntax, bot not in an
arbitrary model.
%and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
The $\eta$ rountrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.

We show another isomorphism between terms depending on a variable and
terms depending on the lifted version of the same variable:
\[
(\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
\]
To derive the roundtrips, we again need admissible equations. Here is
the proof of the first roundtrip equality, the other is analogous:
\begin{alignat*}{10}
  & t[\p^+][\langle\un\,\q\rangle][\p^+][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
  & t[\p^+][{\p^+}^+][\langle\un\,\q[\p^+]\rangle][\langle\mk\,\q\rangle] && {=}(\un[]) \\
  & t[\p^+][{\p^+}^+][\langle\un\,(\q[\p^+])\rangle][\langle\mk\,\q\rangle] && {=}(\q[^+]) \\
  & t[\p^+][{\p^+}^+][\langle\un\,\q\rangle][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,\q[\langle\mk\,\q\rangle]\rangle] && {=}(\un[]) \\
  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\q[\langle\mk\,\q\rangle])\rangle] && {=}(\q[\langle\rangle]) \\
  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\mk\,\q)\rangle] && {=}(\Lift\beta) \\
  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted1}) \\
  & t[\p^+][\p^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted2}) \\
  & t[\p^+][\langle\q\rangle] && {=}(\ref{eq:lifted4}) \\
  & t
\end{alignat*}
Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) && {\cong}\text{($\Lift$ iso)} \\
  & \Tm\,\Gamma\,(\Pi\,A\,B) && {\cong}\text{($\Pi$ iso)} \\
  & \Tm\,(\Gamma\ext A)\,B && {\cong}\text{($\Lift$ iso)} \\
  & \Tm\,(\Gamma\ext A)\,(\Lift\,B) && {\cong}\text{(lifted variable iso)} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) && {\cong}\text{($\Pi$ iso)} \\
  & \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
\end{alignat*}
We hope that the above examples demonstrated the for equations
(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
section.

@@

As a sanity check for our notion of SSC-model, we defined the standard
(metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
SSC-based calculus in Agda. We used the techniques of Kovács
\cite{DBLP:conf/csl/Kovacs22} to define inductive-recursive universes
to model the infinite hierarchy of types, and all the SSC-equations
hold by reflexivity in this model. We defined the analogous CwF-based
standard model.

@@

In this section we show that our SSC-based syntax is isomorphic to the
CwF-based syntax with the same type formers.
\footnote{Using the two translations from SOGATs to GATs
\cite{DBLP:conf/fscd/KaposiX24}, we can rephrase the contents of this
section as follows: the syntaxes of the single and parallel
substitution first order versions of the SOGAT \ref{eq:tt} are
isomorphic.}

@@

The following four equations correspond to equations in B-systems which cannot
be derived from the equations of SSC, these are versions of equations in Section
\ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
\begin{alignat}{10}
  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \label{eq:lifted1}\\
  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2}\\
  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3}\\
  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B\label{eq:lifted4}
\end{alignat}
The corresponding equations on terms are also needed. Note that we do not have
the non-lifted term versions of the last two equations in SSC.

To formally state equations (\ref{eq:lifted1})--(\ref{eq:lifted4}), we define telescopes as an
inductive type together with a recursive operation to append telescopes to
contexts.
\begin{alignat*}{10}
  & \Tel && : \Con\ra\Set \\
  & \blank+\blank && : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con \\
  & \diamond && : \Tel\,\Gamma \\
  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Omega
\end{alignat*}
The appending operation is defined recursively on telescopes.
\begin{alignat*}{10}
  & \Gamma+\diamond && := \Gamma \\
  & \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
\end{alignat*}
We then define a lifting operation over any telescope mutually with an
instantiation operation on telescopes by recursion on telescopes.
\begin{alignat*}{10}
  & \blank[\blank] && : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \\
  & \blank^{+^\Omega} && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
\end{alignat*}

If we try to prove the equations in (\ref{eq:lifted1})--(\ref{eq:lifted4}) by induction on the types
and terms, we run into difficulties in the instantiation case, as we would have
to commute the instantiations without the equations to do so. Therefore, we
first $\alpha$-normalize the syntax to compute away all the instantiations, then
prove the equations by induction on the $\alpha$-normal forms.

@@

Instead of doing induction on $\alpha$-normal forms for each of equations 
(\ref{eq:lifted1})--(\ref{eq:lifted4}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules are
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if $A[\gamma_0] =
  A[\gamma_1]$ and $x[\gamma_0] = x[\gamma_1]$ for any $A:\Ty\,\Gamma$ and
  variable $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\gamma_0^{+^\Omega}] =
  x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}
